---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# butterfly

<!-- badges: start -->
<!-- badges: end -->

The goal of butterfly is to aid in the QA/QC of continually updating/overwritten time-series data where we expect new values over time, but where we want to ensure previous data remains unchanged. Data previously recorded or calculated might change due equipment recalibration, discovery of human error in model code or a change in methodology.

This could have unintended consequences, as changes to previous input data may also alter future predictions in forecasting models.

The butterfly package aims to flag changes to previous data to prevent data changes going unnoticed.

## Installation

You can install the development version of butterfly from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("thomaszwagerman/butterfly")
```

## Example

This is a basic example which shows you how to use butterfly:

```{r simple_example}
library(butterfly)

# Imagine a continually updated dataset, say once a month
jan <- data.frame(
  time = c("2024-01-01", "2023-12-01", "2023-11-01"),
  value = c(0.45, 0.33, 0.24)
)

# In February an additional row appears, all previous data remains the same
feb <- data.frame(
  time = c("2024-02-01", "2024-01-01", "2023-12-01", "2023-11-01"),
  value = c(1.75, 0.45, 0.33, 0.24)
)

# In March an additional row appears again
# ...but a previous value has unexpectedly changed
mar <- data.frame(
  time = c("2024-03-01", "2024-02-01", "2024-01-01", "2023-12-01", "2023-11-01"),
  value = c(2.22, 1.75, 0.45, 1.33, 0.24)
)
```

We can use `butterfly()` to check if our previous values have changed.

```{r butterfly_example}
# Let's use butterfly() to check if our previous values have changed
# And if so, where this change occurred
butterfly(
  feb,
  jan,
  datetime_variable = "time"
)

butterfly(
  mar,
  feb,
  datetime_variable = "time"
)
```

`butterfly()` uses `dplyr::semi_join()` to match the timesteps of your current dataframe, to the timesteps already present in the previous dataframe. `waldo::compare()` is then used to compare these and return the differences.

`butterfly()` follows the `waldo` philosophy of erring on the side of providing too much information, rather than too little. It will give a detailed feedback message on the status between two objects.

### Using butterfly for data wrangling
You might want to return changed rows as a dataframe, or drop them altogether. For this `butterfly_catch()` and `butterfly_release()` are provided.

Here, `butterfly_release()` drops all rows which had changed from the previous version. Note it retains new rows, as these were expected.

```{r butterfly_release}
df_released <- butterfly_release(
  mar,
  feb,
  datetime_variable = "time"
)

df_released
```

## Relevant packages and functions
The butterfly package was created for a specific use case of handling continuously updating/overwritten time-series data, where previous values may change without notice. 

There are other R packages and functions which handle object comparison, which may suit your specific needs better:

* [waldo](https://github.com/r-lib/waldo)
* [diffdf](https://github.com/gowerc/diffdf)

Other functions include `all.equal()` or [dplyr](https://github.com/tidyverse/dplyr)'s `setdiff()`
