---
title: 'Using butterfly in a data processing pipeline'
tags:
  - R
  - quality assurance
  - timeseries
  - ERA5
authors:
  - name: Thomas Zwagerman
    orcid: 0000-0000-0000-0000
    affiliation: 1 
affiliations:
 - name: British Antarctic Survey, UK
   index: 1
date: 23 October 2024
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(butterfly)
```


In this article, we provide a simplified demonstration of butterfly being used in a data processing pipeline.


# Demonstration of Butterfly in a data processing pipeline

You can inspect the full source code of this pipeline in the repository [asli-pipeline](https://github.com/antarctica/asli-pipeline).

## Citations

Andersson, T., & Hosking, J. (2021). Forecasts, neural networks, and results from the paper: 'Seasonal Arctic sea ice forecasting with probabilistic deep learning' (Version 1.0) [Data set]. NERC EDS UK Polar Data Centre. <https://doi.org/10.5285/71820e7d-c628-4e32-969f-464b7efb187c>

Andersson, T.R., Hosking, J.S., Pérez-Ortiz, M. *et al.* Seasonal Arctic sea ice forecasting with probabilistic deep learning. *Nat Commun* **12**, 5124 (2021). <https://doi.org/10.1038/s41467-021-25257-4>

Blair, Gordon S., and Peter A. Henrys. 2023. “The Role of Data Science in Environmental Digital Twins: In Praise of the Arrows.” Environmetrics 34 (January): Not available. https://doi.org/10.1002/env.2789. 

Hersbach, H., Bell, B., Berrisford, P., Biavati, G., Horányi, A., Muñoz Sabater, J., Nicolas, J., Peubey, C., Radu, R., Rozum, I., Schepers, D., Simmons, A., Soci, C., Dee, D., Thépaut, J-N. (2023): ERA5 hourly data on single levels from 1940 to present. Copernicus Climate Change Service (C3S) Climate Data Store (CDS), DOI: 10.24381/cds.adbb2d47

Hosking, J. S., A. Orr, T. J. Bracegirdle, and J. Turner (2016), Future circulation changes off West Antarctica: Sensitivity of the Amundsen Sea Low to projected anthropogenic forcing, Geophys. Res. Lett., 43, 367–376, <doi:10.1002/2015GL067143>.

Tazi, K. (2023). Downscaled ERA5 monthly precipitation data using Multi-Fidelity Gaussian Processes between 1980 and 2012 for the Upper Beas and Sutlej Basins, Himalayas (Version 1.0) [Data set]. NERC EDS UK Polar Data Centre. <https://doi.org/10.5285/b2099787-b57c-44ae-bf42-0d46d9ec87cc>
